---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# iv.sensemakr: Sensitivity Analysis Tools for Instrumental Variables

`iv.sensemakr` implements a suite of sensitivity analysis tools for instrumental variable estimates, as discussed in [Cinelli, C. and Hazlett, C. (2023) "An Omitted Variable Bias Framework for Sensitivity Analysis of Instrumental Variables".](https://carloscinelli.com/files/Cinelli%20and%20Hazlett%20(2020)%20-%20OVB%20for%20IV.pdf) 

##  Development version

To install the development version on GitHub make sure you have the package `devtools` installed.

```{r, eval = FALSE}
# install.packages("devtools") 
devtools::install_github("carloscinelli/iv.sensemakr")
```

# Basic usage

```{r basic-usage, fig.align='center', collapse=T, dpi=400, fig.width=10}
# loads package
library(iv.sensemakr)

# loads dataset
data("card")

# prepares data
y <- card$lwage  # outcome
d <- card$educ   # treatment
z <- card$nearc4 # instrument
x <- model.matrix( ~ exper + expersq + black + south + smsa + reg661 + reg662 +
                     reg663 + reg664 + reg665+ reg666 + reg667 + reg668 + smsa66,
                   data = card) # covariates
# fits IV model
card.fit <- iv_fit(y,d,z,x)

# see results
card.fit

# runs sensitivity analysis
card.sens <- sensemakr(card.fit, benchmark_covariates = c("black", "smsa"))

# see results
card.sens

# sensitivity contour plot
plot(card.sens, lim = 0.09)
```

